Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-03-15 local-capability-machine.tex
changed the standard region to take set instead of range and defined the old one in terms of the new
typos and small corrections

\subsection{Operational Semantics}
Assume a $\decode$ function that decodes words to instructions:
\begin{align*}
  \decode &:\Words \fun \Instrs
\end{align*}
\dominique{mention that it is a simplification to take decode total?}
Assume an $\encodePerm$, $\encodeLoc$, and $\encodePermPair$ function that encodes a permissions, locality, and permission pair, respectively, as an integer:
\begin{align*}
  \encodePerm &: \Perms \fun \ints \\
  \encodeLoc &: \Globals \fun \ints \\
  \encodePermPair &: (\Perms \times \Globals) \fun \ints \\
\end{align*}
Further, assume an inverse function, $\decodePermPair{}$, that decodes permissions
\[
  \decodePermPair{} : \ints \fun (\Perms \times \Globals)
\]

@@

  \sem{\halt}(\Phi)                        & = (\halted,\Phi) \\

@@

  \sem{\load{\refreg{r_1}}{\refheap{r_2}}}  & = 

@@

  \sem{\store{\refheap{r_1}}{\refreg{r_2}}} & = 

@@

  \sem{\move{\refreg{r_1}}{\rv}}            & = \stdUpdatePc{\updateReg{r_1}{\memreg(\rv)}}

@@

  \sem{\lea{\refreg{r_1}}{\rv}}            & =

@@

  \sem{\restricttwo{\refreg{r}}{\rv}}           & =

@@

  \sem{\plus{\refreg{r_1}}{\rv_1}{\rv_2}}               & =

@@

  \sem{\minus{\refreg{r_1}}{\rv_1}{\rv_2}}               & =

@@

  \sem{\subseg{\refreg{r}}{\rv_1}{\rv_2}} & = 

@@

  \sem{\geta{\refreg{r_1}}{\refreg{r_2}}} & = 

@@

  \sem{\getb{\refreg{r_1}}{\refreg{r_2}}} & = 

@@

  \sem{\gete{\refreg{r_1}}{\refreg{r_2}}} & = 

@@

  \sem{\getp{\refreg{r_1}}{\refreg{r_2}}} & = 

@@

  \sem{\getl{\refreg{r_1}}{\refreg{r_2}}} & = 

@@

  \sem{\isptr{\refreg{r}}{\rv}} & =  

@@

              \begin{specification}[Malloc v.2]
                \label{spec:malloc}
                $c_\malloc$ satisfies the specification for malloc iff
                \[  
                  \begin{aligned}
                    &c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
                    &\exists \iota_{\malloc,0} \ldotp \\
                    &\quad (\forall \iota' \futurestr \iota_{\malloc,0} \ldotp \forall W,i \ldotp W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)])) ) \; \land \\
                    &\quad \iota_{\malloc,0}.v = \perma \; \land \\
                    &\quad (\forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
                    &\qquad \forall i, n, \size \in \nats \ldotp \forall
                    w_{\var{ret}} \in \Words \ldotp \\
                    &\qquad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
                    &\qquad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
                    &\qquad \quad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
                    &\qquad \quad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
                    &\qquad \quad \Rightarrow \\
                    &\qquad \qquad\exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
                    &\qquad \qquad \quad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
                    &\qquad \qquad \qquad \Phi \step[j] \Phi' \land \\
                    &\qquad \qquad \qquad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
                    &\qquad \qquad \qquad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
                    &\qquad \qquad \qquad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
                    &\qquad \qquad \qquad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
                    &\qquad \qquad \qquad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
                    &\qquad \qquad \qquad \size - 1 = e'-b' ) \land\\
                    &\quad (\forall \Phi \in \ExecConfs \ldotp (\memreg(r_1)
                    \not\in \ints \vee \memreg(r_1) < 0) \land \memreg(\pcreg) = \updatePcPerm{c_\malloc} \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed)
                  \end{aligned}
                \]
              \end{specification}
              In the specification above $\iota_\malloc'$ is a future region of the initial region that governs malloc.
              \lau{What about $c_\malloc$ and the value relation? Do we add this to the specification or try to prove it based on the specification?}

@@

Standard regions for when writing locally is permitted:
\begin{align*}
  &\iota_{\start,\addrend}^\pwl : \Regions\\
  &\iota_{\start,\addrend}^\pwl \defeq (\temp,1,=,=,H^\pwl_{\start,\addrend}) \\
  \\
  &H^\pwl : \Addrs^2 \fun \States \fun (\Wor \monwknefun \UPred{\HeapSegments})\\
  &H^\pwl_{\start,\addrend}\; s \; \hat{W} \defeq \left\{\npair{\hs} \middle|
    \begin{aligned}
      &\dom(\hs) = [\start,\addrend] \land \\
      &\forall \addr \in [\start,\addrend] \ldotp \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W}))
    \end{aligned}
        \right\} \union \{\npair[0]{\ms} \}
\end{align*}
Revoking all temporary regions:
\begin{align*}
  \revokeTemp{} & : \Worlds \fun \Worlds \\
  \revokeTemp{W} & \defeq \lambda r \ldotp 
                   \begin{cases}
                     \revoked            & \text{if }W(r) = (\temp,s,\phi_\pub,\phi,H) \\
                     W(r)                & \text{otherwise}
                   \end{cases}
\end{align*}
\newcommand{\wrev}[1]{\revokeTemp{#1}}

Standard regions for when write local is not allowed:
\begin{align*}
  &\iota_{\start,\addrend}^{\nwl} : \Regions \\
  &\iota_{\start,\addrend}^{\nwl} \defeq (\temp,1,=,=,H^\nwl_{start,\addrend}) \\
  \\
  &\iota_{\start,\addrend}^{\nwl,p} : \Regions \\
  &\iota_{\start,\addrend}^{\nwl,p} \defeq (\perma,1,=,=,H^\nwl_{start,\addrend})\\
  \\
  &H^\nwl : \Addrs^2 \fun \States \fun (\Wor \monstrnefun \UPred{\HeapSegments})\\
  &H^\nwl_{\start,\addrend} \; s \;\hat{W} \defeq \left\{\npair{\hs} \middle|
    \begin{aligned}
      &\dom(\hs) = [\start,\addrend] \land \\
      &\forall \addr \in [\start,\addrend] \ldotp \\
      &\quad \nonlocal{\ms(\addr)} \land\\ 
      &\quad \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W}))
    \end{aligned}
        \right\} \union \{\npair[0]{\ms} \}
\end{align*}

@@

\subsubsection{LR Sanity lemmas}

@@

\begin{lemma}[Revoke temporary memory with stack]
  \label{lem:revoke-temp-stack}
  \begin{align*}
    & \forall n, \ms, W, \reg, r_\stk,\gl,\start,\addrend,\addr \ldotp\\
    & \quad \memSat{\ms}{W} \land \npair{\reg} \in \stdrr(W) \land \\
    & \quad \reg(r_\stk) = ((\rwlx,\gl),\start,\addrend,\addr) \land b \leq e \\
    & \qquad \exists \ms', \ms_r \ldotp \\
    & \qquad \quad \memSat{\ms }{\revokeTemp{W}} \land \ms = \ms' \uplus \ms_r
  \end{align*}
\end{lemma}
\begin{proof}[lem:revoke-temp-stack]
  % TODO
\end{proof}

\begin{lemma}[Disjoint memory satisfaction]
  \label{lem:disj-mem-sat}
  \begin{align*}
    & \forall n \ldotp \forall \ms, \ms', \ms'' \ldotp \forall W, W', W'' \ldotp\\
    & \quad \ms'' = \ms \uplus \ms' \land W'' = W \uplus W' \land \memSat{\ms}{W} \land \memSat{\ms'}{W'} \Rightarrow \\
    & \qquad \heapSat[\ms \uplus \ms' ]{n}{W \uplus W'}
  \end{align*}
\end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:disj-mem-sat}]
  % TODO
\end{proof}

@@

\begin{lemma}[Data only memory and standard regions]
  \label{lem:mem-sat-data-only-std-regions}
  If
  \begin{itemize}
  \item $\forall \addr \in \dom(\ms)\ldotp \ms(a) \in \nats$
  \item $\iota \in \{\iota^\pwl,\iota^\nwl,\iota^{\nwl,p} \}$
  \end{itemize}
  then
  \[
    \memSat{\ms}{[i \mapsto \iota^\pwl(\dom(\ms))]}
  \]
\end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:mem-sat-data-only-std-regions}]
  % TODO
\end{proof}

>>>>

\subsection{Operational Semantics}
Assume a $\decode$ function that decodes words to instructions:
\begin{align*}
  \decode &:\Words \fun \Instrs
\end{align*}
\dominique{mention that it is a simplification to take decode total?}
Assume an $\encodePerm$, $\encodeLoc$, and $\encodePermPair$ function that encodes a permissions, locality, and permission pair, respectively, as an integer:
\begin{align*}
  \encodePerm &: \Perms \fun \ints \\
  \encodeLoc &: \Globals \fun \ints \\
  \encodePermPair &: (\Perms \times \Globals) \fun \ints \\
\end{align*}
Further, assume a left inverse function, $\decodePermPair{}$, that decodes permissions
\[
  \decodePermPair{} : \ints \fun (\Perms \times \Globals)
\]

@@

  \sem{\halt}(\Phi)                        & = (\halted,\memheap) \\

@@

  \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi)  & = 

@@

  \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) & = 

@@

  \sem{\move{\refreg{r_1}}{\rv}}(\Phi)            & = \stdUpdatePc{\updateReg{r_1}{\memreg(\rv)}}

@@

  \sem{\lea{\refreg{r_1}}{\rv}}(\Phi)            & =

@@

  \sem{\restricttwo{\refreg{r}}{\rv}}(\Phi)           & =

@@

  \sem{\plus{\refreg{r_1}}{\rv_1}{\rv_2}}(\Phi)               & =

@@

  \sem{\minus{\refreg{r_1}}{\rv_1}{\rv_2}}(\Phi)               & =

@@

  \sem{\subseg{\refreg{r}}{\rv_1}{\rv_2}}(\Phi) & = 

@@

  \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) & = 

@@

  \sem{\getb{\refreg{r_1}}{\refreg{r_2}}}(\Phi) & = 

@@

  \sem{\gete{\refreg{r_1}}{\refreg{r_2}}}(\Phi) & = 

@@

  \sem{\getp{\refreg{r_1}}{\refreg{r_2}}}(\Phi) & = 

@@

  \sem{\getl{\refreg{r_1}}{\refreg{r_2}}}(\Phi) & = 

@@

  \sem{\isptr{\refreg{r}}{\rv}}(\Phi) & =  

@@

              \begin{specification}[Malloc Specification]
                \label{spec:malloc}
                $c_\malloc$ satisfies the specification for malloc iff
                \[  
                  \begin{aligned}
                    &c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
                    &\exists \iota_{\malloc,0} \ldotp \\
                    &\quad (\forall \iota' \futurestr \iota_{\malloc,0} \ldotp \forall W,i \ldotp W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)])) ) \; \land \\
                    &\quad \iota_{\malloc,0}.v = \perma \; \land \\
                    &\quad (\forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
                    &\qquad \forall i, n, \size \in \nats \ldotp \forall
                    w_{\var{ret}} \in \Words \ldotp \\
                    &\qquad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
                    &\qquad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
                    &\qquad \quad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
                    &\qquad \quad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
                    &\qquad \quad \Rightarrow \\
                    &\qquad \qquad\exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
                    &\qquad \qquad \quad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
                    &\qquad \qquad \qquad \Phi \step[j] \Phi' \land \\
                    &\qquad \qquad \qquad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
                    &\qquad \qquad \qquad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
                    &\qquad \qquad \qquad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
                    &\qquad \qquad \qquad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
                    &\qquad \qquad \qquad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
                    &\qquad \qquad \qquad \size - 1 = e'-b' ) \land\\
                    &\quad (\forall \Phi \in \ExecConfs \ldotp (\memreg(r_1)
                    \not\in \ints \vee \memreg(r_1) < 0) \land \memreg(\pcreg) = \updatePcPerm{c_\malloc} \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed)
                  \end{aligned}
                \]
              \end{specification}
              In the specification above $\iota_\malloc'$ is a future region of the initial region that governs malloc.
              \lau{What about $c_\malloc$ and the value relation? Do we add this to the specification or try to prove it based on the specification?}

@@

Standard regions for when writing locally is permitted:
\begin{align*}
  &\iota^\pwl : \cal{P} \fun \Regions\\
  &\iota^\pwl\;A \defeq (\temp,1,=,=,H^\pwl \; S) \\
  \\
  &H^\pwl : \cal{P}(\Addrs) \fun \States \fun (\Wor \monwknefun \UPred{\HeapSegments})\\
  &H^\pwl\; A \; s \; \hat{W} \defeq \left\{\npair{\hs} \middle|
    \begin{aligned}
      &\dom(\hs) = A \land \\
      &\forall \addr \in A \ldotp \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W}))
    \end{aligned}
        \right\} \union \{\npair[0]{\ms} \}
\end{align*}
Revoking all temporary regions:
\begin{align*}
  \revokeTemp{} & : \Worlds \fun \Worlds \\
  \revokeTemp{W} & \defeq \lambda r \ldotp 
                   \begin{cases}
                     \revoked            & \text{if }W(r) = (\temp,s,\phi_\pub,\phi,H) \\
                     W(r)                & \text{otherwise}
                   \end{cases}
\end{align*}
Further define
\[
  \iota^\pwl_{\start,\addrend} \defeq \iota^\pwl([\start,\addrend])
\]

\newcommand{\wrev}[1]{\revokeTemp{#1}}

Standard regions for when write local is not allowed:
\begin{align*}
  &\iota^\nwl : \cal{P}(\Addrs) \fun \Regions \\
  &\iota^\nwl\; A \defeq (\temp,1,=,=,H^\nwl\;A) \\
  \\
  &\iota^{\nwl,p} : \cal{P}(\Addrs) \fun \Regions \\
  &\iota^{\nwl,p}\;A \defeq (\perma,1,=,=,H^\nwl\;A)\\
  \\
  &H^\nwl : \cal{P}(\Addrs) \fun \States \fun (\Wor \monstrnefun \UPred{\HeapSegments})\\
  &H^\nwl\;A \; s \;\hat{W} \defeq \left\{\npair{\hs} \middle|
    \begin{aligned}
      &\dom(\hs) = A \land \\
      &\forall \addr \in A \ldotp \\
      &\quad \nonlocal{\ms(\addr)} \land\\ 
      &\quad \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W}))
    \end{aligned}
        \right\} \union \{\npair[0]{\ms} \}
\end{align*}

Further define

\begin{align*}
  \iota^\nwl_{\start,\addrend} & \defeq \iota^\nwl([\start,\addrend])\\
  \iota^{\nwl,p}_{\start,\addrend} & \defeq \iota^{\nwl,p}([\start,\addrend])\\
\end{align*}

@@

\subsubsection{LR Sanity lemmas}
\begin{lemma}
  \label{lem:mem-sat-ne}
  \begin{align*}
    &\forall \hs, n, W \nequal W' \ldotp \\
    &\quad \heapSat[\hs]{n}{W} \land W \nequal W' \Rightarrow \heapSat[\hs]{n}{W'}
  \end{align*}
\end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:mem-sat-ne}]
  %TODO
\end{proof}

@@

%This lemma seems weird given the previous lemmas.
\begin{lemma}[Revoke temporary memory with stack]
  \label{lem:revoke-temp-stack}
  \begin{align*}
    & \forall n, \ms, W, \reg, r_\stk,\gl,\start,\addrend,\addr \ldotp\\
    & \quad \memSat{\ms}{W} \land \npair{\reg} \in \stdrr(W) \land \\
    & \quad \reg(r_\stk) = ((\rwlx,\gl),\start,\addrend,\addr) \land b \leq e \\
    & \qquad \exists \ms', \ms_r \ldotp \\
    & \qquad \quad \memSat{\ms'}{\revokeTemp{W}} \land \ms = \ms' \uplus \ms_r
  \end{align*}
\end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:revoke-temp-stack}]
  % TODO
\end{proof}

\begin{lemma}[Disjoint memory satisfaction]
  \label{lem:disj-mem-sat}
  \begin{align*}
    & \forall n \ldotp \forall \ms, \ms', \ms'' \ldotp \forall W, W', W'' \ldotp\\
    & \quad \ms'' = \ms \uplus \ms' \land W'' = W \uplus W' \land \memSat{\ms}{W} \land \memSat{\ms'}{W'} \Rightarrow \\
    & \qquad \heapSat[\ms'']{n}{W''}
  \end{align*}
\end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:disj-mem-sat}]
  % TODO
\end{proof}

@@

\begin{lemma}[Data only memory and standard regions]
  \label{lem:mem-sat-data-only-std-regions}
  If
  \begin{itemize}
  \item $\forall \addr \in \dom(\ms)\ldotp \ms(a) \in \nats$
  \item $\iota \in \{\iota^\pwl,\iota^\nwl,\iota^{\nwl,p} \}$
  \end{itemize}
  then
  \[
    \memSat{\ms}{[i \mapsto \iota(\dom(\ms))]}
  \]
\end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:mem-sat-data-only-std-regions}]
  % TODO
\end{proof}
